---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# RandomWalker

<!-- badges: start -->
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/RandomWalker)](https://cran.r-project.org/package=RandomWalker)
![](https://cranlogs.r-pkg.org/badges/RandomWalker)
![](https://cranlogs.r-pkg.org/badges/grand-total/RandomWalker)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html##experimental)
[![PRs Welcome](https://img.shields.io/badge/PRs-welcome-brightgreen.svg?style=flat-square)](https://makeapullrequest.com)
<!-- badges: end -->

The goal of RandomWalker is to allow users to easily create Random Walks of different
types that are compatible with the `tidyverse` suite of packages. The package is
currently in the experimental stage of development.

## Installation

You can install the development version of RandomWalker from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("spsanderson/RandomWalker")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(RandomWalker)
## basic example code
rw30()
```
