---
title: "Continuous Distribution Generators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Continuous Distribution Generators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup, echo=FALSE, message=FALSE}
library(RandomWalker)
library(dplyr)
library(ggplot2)
library(patchwork)
```

RandomWalker provides a comprehensive suite of random walk generators based on continuous probability distributions. Each function follows a consistent API design for ease of use.

## Table of Contents

- [Common Parameters](#common-parameters)
- [Normal Distribution](#normal-distribution)
- [Brownian Motion](#brownian-motion)
- [Geometric Brownian Motion](#geometric-brownian-motion)
- [Beta Distribution](#beta-distribution)
- [Cauchy Distribution](#cauchy-distribution)
- [Chi-Squared Distribution](#chi-squared-distribution)
- [Exponential Distribution](#exponential-distribution)
- [F Distribution](#f-distribution)
- [Gamma Distribution](#gamma-distribution)
- [Log-Normal Distribution](#log-normal-distribution)
- [Logistic Distribution](#logistic-distribution)
- [Student's t Distribution](#students-t-distribution)
- [Uniform Distribution](#uniform-distribution)
- [Weibull Distribution](#weibull-distribution)
- [Comparison Guide](#comparison-guide)

## Common Parameters

All continuous distribution generators share these common parameters:

| Parameter | Type | Description | Default |
|-----------|------|-------------|---------|
| `.num_walks` | Integer | Number of walks to generate | 25 |
| `.n` | Integer | Number of steps per walk | 100 |
| `.initial_value` | Numeric | Starting value for the walk | 0 |
| `.dimensions` | Integer | Spatial dimensions (1, 2, or 3) | 1 |

Additional parameters are distribution-specific and control the shape of the probability distribution.

## Normal Distribution

### `random_normal_walk()`

The most commonly used random walk, based on the normal (Gaussian) distribution.

**Function Signature:**
```r
random_normal_walk(
  .num_walks = 25,
  .n = 100,
  .mu = 0,
  .sd = 1,
  .initial_value = 0,
  .dimensions = 1
)
```

**Distribution Parameters:**
- `.mu` - Mean of the distribution (default: 0)
- `.sd` - Standard deviation (default: 1)

**Use Cases:**
- General-purpose random walks
- Modeling measurement errors
- Simulating natural phenomena with normal noise

**Example:**
```{r normal_walk_basic, fig.alt="Line plot showing 10 random walks following a normal distribution, with steps on the x-axis and cumulative values on the y-axis. The walks fluctuate randomly around the starting value."}
# Basic normal walk
random_normal_walk(.num_walks = 10, .n = 100) |>
  visualize_walks()
```

```{r normal_walk_custom, fig.alt="Line plot showing 5 random walks with custom mean and standard deviation parameters. The walks start at 100 and show an upward drift with lower volatility."}
# With custom mean and SD
random_normal_walk(
  .num_walks = 5,
  .n = 200,
  .mu = 0.01,    # Slight upward drift
  .sd = 0.5,     # Lower volatility
  .initial_value = 100
) |> visualize_walks()
```

```{r normal_walk_2d, fig.alt="Scatter plot showing 3 random walks in 2D space, with x-coordinate on the x-axis and y-coordinate on the y-axis. Each walk is represented by connected points forming a path through the 2D plane."}
# 2D spatial walk
random_normal_walk(
  .num_walks = 3,
  .n = 500,
  .dimensions = 2
) |> visualize_walks()
```

### `random_normal_drift_walk()`

Random walk with deterministic drift component.

**Function Signature:**
```r
random_normal_drift_walk(
  .num_walks = 25,
  .n = 100,
  .mu = 0,
  .sd = 1,
  .initial_value = 0,
  .dimensions = 1
)
```

**Key Difference from `random_normal_walk()`:**
- `.drift` - Drift term added to each step (default: 0.1)
- Adds explicit drift term: `drift + random_step`
- More pronounced trending behavior
- Better for modeling processes with clear directional movement

**Example:**
```{r normal_drift_compare, fig.alt="Two panel plot comparing normal walks with and without drift. The top panel shows walks with subtle drift, while the bottom panel shows walks with more pronounced trending behavior due to explicit drift term."}
# Compare walks with and without drift
p1 <- random_normal_walk(.num_walks = 5, .mu = 0.1) |>
  visualize_walks(.pluck = "y") +
  labs(title = "Normal Walk")

p2 <- random_normal_drift_walk(.num_walks = 5, .mu = 0.1) |>
  visualize_walks(.pluck = "y") +
  labs(title = "Normal Walk with Drift")

p1 / p2
```

## Brownian Motion

### `brownian_motion()`

Standard Brownian motion (Wiener process) - the foundation of stochastic calculus.

**Function Signature:**
```r
brownian_motion(
  .num_walks = 25,
  .n = 100,
  .delta_time = 1,
  .initial_value = 0,
  .dimensions = 1
)
```

**Parameters:**
- `.mu` - Drift coefficient (default: 0)
- `.sigma` - Volatility coefficient (default: 1)

**Mathematical Form:**
```
dX(t) = μ dt + σ dW(t)
```

**Use Cases:**
- Financial mathematics (Black-Scholes model)
- Physics (particle diffusion)
- Signal processing (noise modeling)

**Example:**
```{r brownian_basic, fig.alt="Line plot showing 10 standard Brownian motion paths. The paths start at zero and fluctuate randomly with continuous movements typical of Wiener processes."}
# Standard Brownian motion
brownian_motion(.num_walks = 10) |>
  visualize_walks()
```

```{r brownian_drift_vol, fig.alt="Line plot showing 50 Brownian motion paths with custom drift and volatility parameters. The paths start at 100 and show semi-transparent overlapping trajectories."}
# With drift and volatility
brownian_motion(
  .num_walks = 50,
  .n = 252,
  .delta_time = 0.05,
  .initial_value = 100
) |> visualize_walks(.alpha = 0.3)
```

### `geometric_brownian_motion()`

Geometric Brownian motion - the standard model for stock prices.

**Function Signature:**
```r
geometric_brownian_motion(
  .num_walks = 25,
  .n = 100,
  .mu = 0,
  .sigma = 0.1,
  .initial_value = 100,
  .delta_time = 0.003,
  .dimensions = 1
)
```

**Parameters:**
- `.mu` - Expected return (drift)
- `.sigma` - Volatility

**Mathematical Form:**
```
dS(t) = μ S(t) dt + σ S(t) dW(t)
S(t) = S(0) exp((μ - σ²/2)t + σW(t))
```

**Key Properties:**
- Always positive (can't go below zero)
- Log-normal distribution of prices
- Percentage changes are normally distributed

**Use Cases:**
- Stock price modeling
- Option pricing
- Asset allocation simulations
- Monte Carlo risk analysis

**Example:**
```{r gbm_stock_sim, fig.alt="Line plot showing 1000 semi-transparent stock price simulation paths using geometric Brownian motion. All paths start at 100 and spread out over time, remaining positive throughout."}
# Model stock prices
stock_sim <- geometric_brownian_motion(
  .num_walks = 1000,
  .n = 252,        # Trading days in a year
  .mu = 0.08,      # 8% expected return
  .sigma = 0.25,   # 25% volatility
  .initial_value = 100
)

# Visualize scenarios
stock_sim |> visualize_walks(.alpha = 0.1)
```

```{r gbm_analysis, eval=FALSE}
# Analyze outcomes
stock_sim |>
  summarize_walks(.value = cum_prod_y, .group_var = walk_number) |>
  summarize(
    median_price = median(max_val),
    prob_profit = mean(max_val > 100),
    prob_loss_50 = mean(min_val < 50)
  )
```

## Beta Distribution

### `random_beta_walk()`

Random walk based on the beta distribution (bounded between 0 and 1).

**Function Signature:**
```r
random_beta_walk(
  .num_walks = 25,
  .n = 100,
  .shape1 = 2,
  .shape2 = 2,
  .ncp = 0,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.shape1` - First shape parameter (α)
- `.shape2` - Second shape parameter (β)
- `.ncp` - Non-centrality parameter (default: 0)

**Properties:**
- Steps are bounded: 0 ≤ step ≤ 1
- Flexible shapes (uniform, U-shaped, bell-shaped)
- Mean = α / (α + β)
- Variance = (α β) / ((α + β)² (α + β + 1))

**Shape Guide:**
- `shape1 = 1, shape2 = 1`: Uniform (0,1)
- `shape1 = 2, shape2 = 2`: Symmetric, bell-shaped
- `shape1 = 2, shape2 = 5`: Right-skewed
- `shape1 = 5, shape2 = 2`: Left-skewed

**Use Cases:**
- Modeling proportions or percentages
- Bounded processes (e.g., utilization rates)
- Success rates over time

**Example:**
```{r beta_symmetric, fig.alt="Line plot showing 10 random walks following a symmetric beta distribution with shape parameters 2 and 2. The walks fluctuate with steps bounded between 0 and 1."}
# Symmetric beta walk
random_beta_walk(
  .num_walks = 10,
  .shape1 = 2,
  .shape2 = 2
) |> visualize_walks()
```

```{r beta_skewed, fig.alt="Line plot showing 10 random walks following a right-skewed beta distribution with shape parameters 2 and 5. The walks show a tendency toward lower values."}
# Right-skewed (toward 0)
random_beta_walk(
  .num_walks = 10,
  .shape1 = 2,
  .shape2 = 5
) |> visualize_walks()
```

```{r beta_compare, fig.alt="Two panel plot comparing beta distributions with different shape parameters. The top panel shows uniform (1,1) beta walks, while the bottom panel shows right-skewed (2,5) beta walks."}
# Compare different shapes
p1 <- random_beta_walk(.shape1 = 1, .shape2 = 1) |>
  visualize_walks(.pluck = "y")
p2 <- random_beta_walk(.shape1 = 2, .shape2 = 5) |>
  visualize_walks(.pluck = "y")
p1 / p2
```

## Cauchy Distribution

### `random_cauchy_walk()`

Random walk with heavy tails - extreme values are much more common than in normal distribution.

**Function Signature:**
```r
random_cauchy_walk(
  .num_walks = 25,
  .n = 100,
  .location = 0,
  .scale = 1,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.location` - Location parameter (median)
- `.scale` - Scale parameter (spread)

**Properties:**
- **No defined mean or variance!**
- Heavy tails (high probability of extreme values)
- Symmetric around location
- Much more volatile than normal distribution

**Use Cases:**
- Modeling extreme events
- Financial crises simulation
- Physics (resonance phenomena)
- Robust statistics demonstrations

**Example:**
```{r cauchy_basic, fig.alt="Line plot showing 10 random walks following a Cauchy distribution. The walks exhibit extreme volatility with large jumps due to the heavy-tailed distribution."}
# Standard Cauchy walk
random_cauchy_walk(.num_walks = 10) |>
  visualize_walks()
```

```{r cauchy_compare, fig.alt="Two panel plot comparing normal walks (top) versus Cauchy walks (bottom). The Cauchy walks show much more extreme movements and larger fluctuations due to heavy tails."}
# Compare with normal walk
p1 <- random_normal_walk(.num_walks = 5, .sd = 1) |>
  visualize_walks(.pluck = "y") +
  labs(title = "Normal Walk")

p2 <- random_cauchy_walk(.num_walks = 5, .scale = 1) |>
  visualize_walks(.pluck = "y") +
  labs(title = "Cauchy Walk (Heavy Tails)")

p1 / p2
```

## Chi-Squared Distribution

### `random_chisquared_walk()`

Random walk based on chi-squared distribution (always positive).

**Function Signature:**
```r
random_chisquared_walk(
  .num_walks = 25,
  .n = 100,
  .df = 5,
  .ncp = 0,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.df` - Degrees of freedom
- `.ncp` - Non-centrality parameter (default: 0)

**Properties:**
- Steps are always positive
- Right-skewed (especially for low df)
- Mean = df
- Variance = 2df

**Use Cases:**
- Goodness-of-fit tests
- Variance estimation
- Quality control
- Reliability engineering

**Example:**
```{r chisq_low_df, fig.alt="Line plot showing 10 random walks following a chi-squared distribution with 1 degree of freedom. The walks are strongly right-skewed with always positive steps."}
# Low df (very skewed)
random_chisquared_walk(.num_walks = 10, .df = 1) |>
  visualize_walks()
```

```{r chisq_high_df, fig.alt="Line plot showing 10 random walks following a chi-squared distribution with 10 degrees of freedom. The walks are more symmetric than the low df case but still show positive steps only."}
# Higher df (more symmetric)
random_chisquared_walk(.num_walks = 10, .df = 10) |>
  visualize_walks()
```

## Exponential Distribution

### `random_exponential_walk()`

Random walk with exponentially distributed steps (memoryless property).

**Function Signature:**
```r
random_exponential_walk(
  .num_walks = 25,
  .n = 100,
  .rate = 1,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.rate` - Rate parameter (λ)

**Properties:**
- Steps are always positive
- Memoryless property
- Mean = 1/rate
- Variance = 1/rate²

**Use Cases:**
- Time between events (queuing theory)
- Lifetime modeling
- Radioactive decay
- Poisson process intervals

**Example:**
```{r exp_basic, fig.alt="Line plot showing 10 random walks following a standard exponential distribution. The walks increase monotonically with always positive steps."}
# Standard exponential
random_exponential_walk(.num_walks = 10) |>
  visualize_walks()
```

```{r exp_fast, fig.alt="Line plot showing 10 random walks with fast exponential rate (5). The walks show smaller steps compared to the standard rate."}
# Fast rate (smaller steps)
random_exponential_walk(.num_walks = 10, .rate = 5) |>
  visualize_walks()
```

```{r exp_slow, fig.alt="Line plot showing 10 random walks with slow exponential rate (0.5). The walks show larger steps compared to the standard rate."}
# Slow rate (larger steps)
random_exponential_walk(.num_walks = 10, .rate = 0.5) |>
  visualize_walks()
```

## F Distribution

### `random_f_walk()`

Random walk based on F-distribution (ratio of chi-squared variables).

**Function Signature:**
```r
random_f_walk(
  .num_walks = 25,
  .n = 100,
  .df1 = 5,
  .df2 = 5,
  .ncp = NULL,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.df1` - Numerator degrees of freedom
- `.df2` - Denominator degrees of freedom
- `.ncp` - Non-centrality parameter

**Use Cases:**
- ANOVA
- Variance ratio tests
- Model comparison

**Example:**
```{r f_dist, fig.alt="Line plot showing 10 random walks following an F-distribution with 5 and 10 degrees of freedom. The walks show right-skewed behavior typical of F-distributions."}
random_f_walk(.num_walks = 10, .df1 = 5, .df2 = 10) |>
  visualize_walks()
```

## Gamma Distribution

### `random_gamma_walk()`

Flexible distribution for positive values.

**Function Signature:**
```r
random_gamma_walk(
  .num_walks = 25,
  .n = 100,
  .shape = 1,
  .scale = 1,
  .rate = NULL,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.shape` - Shape parameter (k)
- `.rate` - Rate parameter (β)
- `.scale` - Scale parameter (1/rate)

**Properties:**
- Steps always positive
- Flexible shapes
- Mean = shape/rate
- Variance = shape/rate²

**Use Cases:**
- Waiting times
- Rainfall modeling
- Insurance claims
- Reliability analysis

**Example:**
```{r gamma_shape1, fig.alt="Line plot showing 10 random walks following a gamma distribution with shape parameter 1. The walks show exponential-like behavior with always positive steps."}
# Different shape parameters
random_gamma_walk(.num_walks = 10, .shape = 1, .rate = 1) |>
  visualize_walks()
```

```{r gamma_shape5, fig.alt="Line plot showing 10 random walks following a gamma distribution with shape parameter 5. The walks show more symmetric behavior compared to shape 1."}
random_gamma_walk(.num_walks = 10, .shape = 5, .rate = 1) |>
  visualize_walks()
```

## Log-Normal Distribution

### `random_lognormal_walk()`

Random walk where log of steps is normally distributed.

**Function Signature:**
```r
random_lognormal_walk(
  .num_walks = 25,
  .n = 100,
  .meanlog = 0,
  .sdlog = 1,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.meanlog` - Mean of log
- `.sdlog` - Standard deviation of log

**Properties:**
- Steps always positive
- Right-skewed
- Multiplicative processes

**Use Cases:**
- Income distributions
- File sizes
- Particle sizes
- Stock prices (alternative to GBM)

**Example:**
```{r lognormal, fig.alt="Line plot showing 10 random walks following a log-normal distribution. The walks are right-skewed with always positive steps typical of log-normal distributions."}
random_lognormal_walk(.num_walks = 10) |>
  visualize_walks()
```

## Logistic Distribution

### `random_logistic_walk()`

Random walk with logistic distribution (heavier tails than normal).

**Function Signature:**
```r
random_logistic_walk(
  .num_walks = 25,
  .n = 100,
  .location = 0,
  .scale = 1,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Use Cases:**
- Growth models
- Binary classification
- Neural networks

**Example:**
```{r logistic, fig.alt="Line plot showing 10 random walks following a logistic distribution. The walks have symmetric fluctuations with slightly heavier tails than normal distribution."}
random_logistic_walk(.num_walks = 10) |>
  visualize_walks()
```

## Student's t Distribution

### `random_t_walk()`

Random walk with t-distribution (adjustable tail heaviness).

**Function Signature:**
```r
random_t_walk(
  .num_walks = 25,
  .n = 100,
  .df = 5,
  .initial_value = 0,
  .ncp = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.df` - Degrees of freedom (controls tail heaviness)
- `.ncp` - Non-centrality parameter

**Properties:**
- Heavier tails than normal (for low df)
- Approaches normal as df → ∞
- Mean = 0 (if df > 1)
- Variance = df/(df-2) (if df > 2)

**Use Cases:**
- Robust statistics
- Small sample inference
- Financial returns modeling

**Example:**
```{r t_heavy, fig.alt="Line plot showing 10 random walks following a t-distribution with 3 degrees of freedom. The walks show heavy tails with occasional large jumps."}
# Heavy tails (df = 3)
random_t_walk(.num_walks = 10, .df = 3) |>
  visualize_walks()
```

```{r t_normal, fig.alt="Line plot showing 10 random walks following a t-distribution with 30 degrees of freedom. The walks closely resemble normal walks with lighter tails."}
# Nearly normal (df = 30)
random_t_walk(.num_walks = 10, .df = 30) |>
  visualize_walks()
```

## Uniform Distribution

### `random_uniform_walk()`

Random walk with uniformly distributed steps.

**Function Signature:**
```r
random_uniform_walk(
  .num_walks = 25,
  .n = 100,
  .min = 0,
  .max = 1,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.min` - Minimum value
- `.max` - Maximum value

**Properties:**
- All values equally likely
- Mean = (min + max) / 2
- Variance = (max - min)² / 12

**Use Cases:**
- Modeling equal probabilities
- Random sampling
- Monte Carlo simulations

**Example:**
```{r uniform, fig.alt="Line plot showing 10 random walks with uniformly distributed steps between -1 and 1. The walks show bounded steps with all values equally likely."}
# Symmetric around 0
random_uniform_walk(.num_walks = 10, .min = -1, .max = 1) |>
  visualize_walks()
```

## Weibull Distribution

### `random_weibull_walk()`

Random walk with Weibull distribution (flexible lifetime model).

**Function Signature:**
```r
random_weibull_walk(
  .num_walks = 25,
  .n = 100,
  .shape = 1,
  .scale = 1,
  .initial_value = 0,
  .samp = TRUE,
  .replace = TRUE,
  .sample_size = 0.8,
  .dimensions = 1
)
```

**Parameters:**
- `.shape` - Shape parameter (k)
- `.scale` - Scale parameter (λ)

**Properties:**
- Steps always positive
- shape < 1: Decreasing failure rate
- shape = 1: Constant failure rate (exponential)
- shape > 1: Increasing failure rate

**Use Cases:**
- Reliability engineering
- Survival analysis
- Wind speed modeling
- Materials science

**Example:**
```{r weibull_low, fig.alt="Line plot showing 10 random walks following a Weibull distribution with shape parameter 0.5. The walks show decreasing failure rate behavior with always positive steps."}
# Different shape parameters
random_weibull_walk(.num_walks = 10, .shape = 0.5) |>
  visualize_walks()
```

```{r weibull_high, fig.alt="Line plot showing 10 random walks following a Weibull distribution with shape parameter 2. The walks show increasing failure rate behavior."}
random_weibull_walk(.num_walks = 10, .shape = 2) |>
  visualize_walks()
```

## Comparison Guide

### When to Use Each Distribution

| Distribution | Use When... | Key Property |
|--------------|-------------|--------------|
| Normal | General modeling, natural phenomena | Symmetric, bell-shaped |
| Brownian Motion | Continuous stochastic processes | Foundation of stochastic calculus |
| Geometric Brownian | Stock prices, always positive | Log-normal, multiplicative |
| Beta | Bounded processes (0-1) | Flexible shapes, bounded |
| Cauchy | Extreme events, heavy tails | No mean/variance |
| Chi-Squared | Sum of squares, variance tests | Right-skewed, positive |
| Exponential | Time between events | Memoryless property |
| F | Variance ratios | Ratio of chi-squared |
| Gamma | Positive values, waiting times | Flexible, positive |
| Log-Normal | Multiplicative processes | Right-skewed, positive |
| Logistic | S-curves, classification | Heavier tails than normal |
| Student's t | Robust modeling, small samples | Adjustable tail heaviness |
| Uniform | Equal probabilities | All values equally likely |
| Weibull | Reliability, survival | Flexible failure rates |

### Tail Behavior

**Light Tails** (fewer extreme values):
- Uniform

**Medium Tails**:
- Normal
- Exponential
- Gamma (shape > 1)

**Heavy Tails** (more extreme values):
- Logistic
- Student's t (low df)
- Cauchy (extreme)

### Symmetry

**Symmetric**:
- Normal
- Brownian Motion
- Cauchy
- Logistic
- Student's t
- Uniform

**Right-Skewed** (tail extends right):
- Beta (depends on parameters)
- Chi-Squared
- Exponential
- F
- Gamma
- Log-Normal
- Weibull (depends on shape)

## Next Steps

- **Discrete Distribution Generators** - Explore discrete distributions
- **Multi-Dimensional Walks** - Work in 2D and 3D
- **Visualization Guide** - Create beautiful plots
- **Use Cases and Examples** - Real-world applications

---

**Need help choosing?** Check out the **FAQ** or **Use Cases** for guidance!
